\section{\LangA}

% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
\subsection{Functions}

\begin{program}
term identity (α:Type) (a:α) : α : a.

term compose (α β γ : Type)
  : (α → β) → (β → γ) → (α → γ)
  ≔ f g a ⇒ f (g a).
\end{program}

\begin{notation}[caption={Notation for \code|compose|}]
$\mvar{term}_1$ ○ $\mvar{term}_2$   $\syneq$   compose $\mvar{term}_1$ $\mvar{term}_2$
\end{notation}

% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
\subsection{Data}

\begin{program}[caption={unit}]
primitive type unit.

// only constructor
primitive term • : unit.
\end{program}


\begin{program}[caption={boolean}]
type boolean : Type { true | false }.
\end{program}

\begin{notational}[caption={Notation for conditional}]
if $\mvar{term}_1$ then $\mvar{term}_2$ else $\mvar{term}_3$
  $\syneq$
    cases $\mvar{term}_1$ { true  ⇒ $\mvar{term}_2$
                          | false ⇒ $\mvar{term}_3$
\end{notational}


\begin{program}[caption={natural number}]
type natural : Type
  { zero : natural
  | succ : natural → natural }.

// canonical terms
term N0 : natural ≔ zero.
term N1 : natural ≔ succ N0.
term N2 : natural ≔ succ N1.
term N3 : natural ≔ succ N2.
// $\cdots$ and so on

// addition
term add (m n : natural) : natural
  ≔ cases m
      { zero    ⇒ n
      | succ m' ⇒ succ (add m' n) }.

// multiplication
term mul (m n : natural) : natural
  ≔ cases m
      { zero    ⇒ zero
      | succ m' ⇒ add n (mul m' n) }.

// minimum
term min (m n : natural) : natural
  ≔ cases m
      { zero    ⇒ zero
      | succ m' ⇒ cases n
                    { zero    ⇒ m'
                    | succ n' ⇒ succ (min m' n') } }.

// maximum
term max (m n : natural) : natural
  ≔ cases m
      { zero    ⇒ n
      | succ m' ⇒ cases n
                    { zero    ⇒ m
                    | succ n' ⇒ succ (max m' n') } }.
\end{program}

\begin{program}[caption={integer}]
type integer : Type
  { negative : natural → integer
  | positive : natural → integer }.

// canonical terms
basic term  0 : integer ≔ positive N0.
basic term  1 : integer ≔ positive N1
basic term -1 : integer ≔ negative N1
basic term  1 : integer ≔ positive N2
basic term -1 : integer ≔ negative N2
// $\cdots$ and so on

// negation
term neg (i : integer) : integer
  ≔ cases i
      { left  n ⇒ right n
      ; right n ⇒ left  n }.

// addition; inherits notation
term add (i j : integer) : integer ≔ ...

// subtraction; inherits notation
term sub (i j : integer) : integer ≔ ...

// modulus
term mod (i j : integer) : integer
  ≔ if i < j
      then i
      els mod (i - j) j.
\end{program}

\begin{program}[caption={rational}]
type rational : Type
  { numerator   : integer
  ; denomenator : natural }.

basic term 0   : rational ≔ (0, 1).
basic term 1   : rational ≔ (1, 1).
basic term 1/2 : rational ≔ (1, 2).
basic term 2/8 : rational ≔ (2, 8).

term div (q r : rational) : rational ≔ ...
\end{program}

\begin{notational}[caption={Notations for infixed binary numerical operations.}]
$\mvar{term}_1$ + $\mvar{term}_2$   $\syneq$   add $\mvar{term}_1$ $\mvar{term}_2$

$\mvar{term}_1$ * $\mvar{term}_2$   $\syneq$   mul $\mvar{term}_1$ $\mvar{term}_2$

- $\mvar{term}$   $\syneq$   neg $\mvar{term}$

$\mvar{term}_1$ - $\mvar{term}_2$   $\syneq$   sub $\mvar{term}_1$ $\mvar{term}_2$

$\mvar{term}_1$ / $\mvar{term}_2$   $\syneq$   div $\mvar{term}_1$ $\mvar{term}_2$
\end{notational}

\subparagraph{Characters and Strings.}
%
These data types are introduced primitively because it is much to annoying and irrelevant to formally introduce them.

\begin{program}[caption={string}]
primitive type character.
primitive type string.

// utilities
primitive term character-join     : list character → string.
primitive term string-join        : string → string → string.
primitive term string-concat      : list string → string.
primitive term string-explode     : string → list character.
primitive term character-append   : character → string → string.
primitive term character-unappend : string → character ⊗ string.
\end{program}

\begin{notational}[caption={Notations for characters and strings.}]
$\mvar{term}_1$ ⧺ $\mvar{term}_2$   $\syneq$   string-concat [$\mvar{term}_1$, $\mvar{term}_2$]
\end{notational}

% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
% ------------------------------------------------------------------------------
\subsection{Data Structures}

\begin{program}[caption={optional}]
type optional (α : Type)
  { some : α → optional α
  | none : optioanl α }.
\end{program}


\begin{program}[caption={list}]
type list (α : Type) : Type ≔ unit ⊕ (α ✕ list α).

type list (α : Type) : Type
  { nil : list α
  | app : α → list α → list α }.
\end{program}

\begin{notational}[caption={list notations}]
[]   $\syneq$   nil

$\mvar{term}_1$ ∷ $\mvar{term}_2$   $\syneq$   app $\mvar{term}_1$ $\mvar{term}_2$

$\mvar{term}_1$ ◇ $\mvar{term}_2$   $\syneq$   concat $\mvar{term}_1$ $\mvar{term}_2$
\end{notational}

\begin{program}[caption={list utilities}]
term concat (α : Type) (ass : list (list α)) : list α
  ≔ cases ass
      { []        ⇒ []
      | as ∷ ass' ⇒ cases as
                      { []      ⇒ concat ass'
                      | a ∷ as' ⇒ a ∷ concat (as' ∷ ass') }.

// undefined on nil
term head (α : Type) (as : list α) : α
  ≔ cases as{ a ∷ _ ⇒ a }.

// tail of empty is empty
term tail (α : Type) (as : list α) : list α
  ≔ cases as{ [] ⇒ [] | _ ∷ as' ⇒ as' }.

// undefined on out-of-bound indices
// gets the $n$th element of $as$
term index (α : Type) (as : list α) (n : natural) : α
  ≔ cases n
      { zero    ⇒ cases as{ a ∷ _   ⇒ a }
      | succ n' ⇒ cases as{ _ ∷ as' ⇒ index as' n' } }.

term contains (α : Type) (as : list α) (a : α) : boolean
  ≔ cases as
      { []       ⇒ false
      | a' ∷ as' ⇒ if a = a' then true else contains as' a }.

term replicate (α : Type) (n : natural) (a : α) : list α
  ≔ cases n
      { zero    ⇒ []
      | succ n' ⇒ a ∷ repeat n' a }.
\end{program}


\begin{program}[caption={tree and forest}]
type tree (α : Type) : Type
  { value    : α
  ; branches : list (tree α) }.

type forest (α : Type) : Type ≔ list (tree α).
\end{program}


\begin{program}[caption={mapping}]
type mapping (α β : Type) : Type ≔ α → β.

term make-mapping (α β : Type) (ls : list (α ✕ β)) : mapping α β
  ≔ $\cdots$

term lookup (α β : Type) (m : mapping α β) (a : α) : β
  ≔ m a.
\end{program}

\begin{notational}[caption={mapping notations}]
$\mvar{type}_1$ ⇛ $\mvar{type}_2$   $\syneq$   mapping $\mvar{type}_1$ $\mvar{type}_2$.

$〚$ $\mvar{term}_{i1}$ ↦ $\mvar{term}_{i2}$ $〛$   $\syneq$   make-mapping [$〚$ ($\mvar{term}_{i1}$, $\mvar{term}_{i2}$) $〛$]

$\mvar{term}_1$@$\mvar{term}_2$   $\syneq$   lookup $\mvar{term}_1$ $\mvar{term}_2$.
\end{notational}
